Refinement of Thermostated Molecular Dynamics Using Backward Error Analysis

### 2019-02-26 response_to_reviewers.tex

This is a nicely presented article that is likely to be of interest the readership of JCP. Overall, it is technically sound, but I have a couple of comments that the authors should consider. 
}}

\textbf{Answer}:
 We sincerely thank the reviewer for the kind comments and helpful suggestions. They have been considered in full. Please find our answers below.

{\color{blue}{
(1) There is a recent paper by Smit and co-workers on equipartitioning in thermostatted MD simulations [Smit and co-workers J. Chem. Theory Comput., 2018, 14 (10), pp 5262-5272]. There is no direct overlap with the work presented here, but the authors may wish to add this reference for interested readers and discuss the results in the context of the current study. 
}}

\textbf{Answer}:
We thank the reviewer for pointing out this reference.
A citation has been included in the introductory section (on page 1 of the manuscript).

@@

{\color{blue}{
(2) The authors assess the magnitude of sampling errors by examining the dependence of average thermodynamics properties on time step. This assumes that correct canonical sampling is recovered when using a small time step, but this may not be true in practice for a number of reasons (e.g., errors in thermostat implementation). The authors should consider analyzing the sample distributions rather than just the averages. To this end, they may wish to consider implementing some of the numerical tests described in J. Chem. Theory Comput., 2013, 9 (2), pp 909-926. 
}}

\textbf{Answer}:
The reviewer is right.
We have applied the numerical tests introduced in the suggested reference (Shirts, 2013) and verified that the sampled energies are in accordance with the canonical distribution.
We have included this information on page 6 of the manuscript.
It is important to note that, for time-step sizes with noticeable discretization errors, the numerical tests are not suitable for analyzing the potential and kinetic energies separately.
They assume that the Hamiltonian can be separated into exclusively momentum- and position-dependent parts, which is not true in these cases.
Therefore, we applied the maximum likelihood approach only to the total energies.
The results for the NVT-MD methods are reported in the table below.
Note that the deviations from the true slope are always smaller than 3$\sigma$ for time-step sizes up to 5 fs.
Higher deviations, according to Shirts (2013), would indicate inconsistency with the canonical ensemble (e.g. the author found values above 40$\sigma$ for the knowingly problematic Berendsen thermostat).
Because the numerical values in the table below would not be essential to the discussions in the manuscript, we decided to leave them out.

\begin{table*}[h!]
	\caption{Ensemble validation of the NVT-MD methods using the maximum likelihood approach for different time-step sizes}
	\label{table:check_ens}
		\centering
		\resizebox{\columnwidth}{!}{%
	\begin{tabular}{r c c c c c c}
		{$\timestep$} & {estimated $\Delta T$} & {$\sigma$ deviations}  & {estimated $\Delta T$}  & {$\sigma$ deviations}  &    {estimated $\Delta T$}  & {$\sigma$ deviations} \\
		{(fs)} & {(K)} &   & {(K)}  &  & {(K)}  & \\
		\hline
		&\multicolumn{2}{c}{Refined NHC} & \multicolumn{2}{c}{NHC }   &	\multicolumn{2}{c}{KLN}\\
		1.0	& 3.02(2) & 1.07 & 2.98(2) & 0.66 &  2.97(2) & 1.16 \\
		2.0 & 3.02(2) & 0.81 & 2.98(2) & 0.73 &  3.00(2) & 0.02 \\
		3.0 & 2.98(2) & 0.73 & 2.95(2) & 2.33  & 2.99(4) & 0.18 \\
		4.0 & 3.01(2) & 0.41 & 3.01(2) & 0.65 & 2.92(8) & 0.99 \\
		5.0 & 3.02(2) & 0.79 & 2.93(2) & 3.05  & 3.02(5) & 0.40 \\
		6.0 & 3.03(3) & 1.23  & 2.92(2) & 3.87  & 3.14(5) & 2.69  \\
		7.0 & 2.98(2) & 0.74 & 2.89(2) & 5.35 & 3.03(4) & 0.73 \\
		\hline
	\end{tabular}
}
\end{table*}

### 2019-02-26 rigid_bodies.tex
Resubmitted (revised) version

Molecular dynamics (MD) simulation involves several types of approximations in the quest for mimicking real systems.
Naively, we tend to rank simulation results by their proximity to experimental data and attribute any disagreement to a lack of accuracy of the employed force field.
However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
It has been recently shown \cite{Palmer_2018} that erroneous simulation results \cite{Limmer_2011, Limmer_2013} were the cause of a long-standing controversy \cite{Smart_2018} about a liquid-liquid transition believed to exist for supercooled water.
In this case, hybrid Monte Carlo steps were employed with initial velocities that failed to comply with the equipartition between translational and rotational degrees of freedom \cite{Palmer_2018}.
As a matter of fact, similar ill-partitioning of kinetic energy can arise in MD simulations as an artifact due to the discretization of the equations of motion.
This can affect not only the translation/rotation equipartition in rigid-body systems \cite{Davidchack_2010, Silveira_2017}, but also the distribution of energy among molecules of distinct sizes \cite{Eastwood_2010}.
The issue with rigid-body MD is relevant not only because water \cite{Jorgensen_1983} and other small molecules are usually modeled as rigid bodies, but also because designing proteins, nanoparticles, or other large structures as collections of interconnected bodies \cite{Miller_2002, Knorowski_2012, Patra_2013} can be seen as a useful coarse-graining strategy.
It also brings into question the NVT-MD method developed by Kamberaj, Low, and Neal \cite{Kamberaj_2005} (KLN) who, by assuming equipartition, coupled two independent thermostats to the translational and rotational degrees of freedom.
As we will show here, the method fails to reproduce the specified temperature even for time-step sizes typically employed in MD.
In fact, Davidchack \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
Nonetheless, the cited double-thermostat method is available in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995} and has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}, in spite of all the issues observed here.

@@

Although the results are not shown here, we have applied a method, introduced and implemented by Shirts ~\cite{Shirts_2013}, that serves for checking the consistency between the sampled energy distributions and the theoretical ones. For the NVT-MD methods investigated, we have verified that, generally, the total energy distributions deviate 1$\sigma$ or less from the theoretical ones for time-step sizes up to 5 fs.

>>>>

Molecular dynamics (MD) simulation involves several types of approximations in the quest for mimicking real systems.
Naively, we tend to rank simulation results by their proximity to experimental data and attribute any disagreement to a lack of accuracy of the employed force field.
However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
It has been recently shown \cite{Palmer_2018} that erroneous simulation results \cite{Limmer_2011, Limmer_2013} were the cause of a long-standing controversy \cite{Smart_2018} about a liquid-liquid transition believed to exist for supercooled water.
In this case, hybrid Monte Carlo steps were employed with initial velocities that failed to comply with the equipartition between translational and rotational degrees of freedom \cite{Palmer_2018}.
As a matter of fact, similar ill-partitioning of kinetic energy can arise in MD simulations, either associated with noncanonical rescaling of velocities \cite{Braun_2018} or as an artifact due to the discretization of the equations of motion.
This can affect not only the translation/rotation equipartition in rigid-body systems \cite{Davidchack_2010, Silveira_2017}, but also the distribution of energy among molecules of distinct sizes \cite{Eastwood_2010}.
The issue with rigid-body MD is relevant not only because water \cite{Jorgensen_1983} and other small molecules are usually modeled as rigid bodies, but also because designing proteins, nanoparticles, or other large structures as collections of interconnected bodies \cite{Miller_2002, Knorowski_2012, Patra_2013} can be seen as a useful coarse-graining strategy.
It also brings into question the NVT-MD method developed by Kamberaj, Low, and Neal \cite{Kamberaj_2005} (KLN) who, by assuming equipartition, coupled two independent thermostats to the translational and rotational degrees of freedom.
As we will show here, the method fails to reproduce the specified temperature even for time-step sizes typically employed in MD.
In fact, Davidchack \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
Nonetheless, the cited double-thermostat method is available in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995} and has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}, in spite of all the issues observed here.

@@

In order to verify if the total energies sampled in the simulations were consistent with the canonical ensemble, we employed the testing procedure developed by Shirts in Ref.~\citenum{Shirts_2013}.
All the NVT-MD methods investigated here succeeded in the test for time-step sizes up to 5 fs (results not shown).
